Introduction to *cellcount* - Draft vignette

### 2023-09-06 cellcount_intro.Rmd
Updates to vingette from comments

The ***cellcount*** package provides a simple approach to quantifying bacteria and cyanobacteria *via* flourescence microscopy. Microbial enumeration is an important tool for validating new methods and critical to assess potential ecological changes. However, standard manual microbial enumeration is time-consuming and notoriously difficult due to large colonies, filaments, and dense biomats. Conventional counting methods like microscopy require specialized expertise and are error prone due to variations between microscopists. Additionally, other methods of counting cells or particles (i.e. coulter counter, flow cytometer, etc.) are not suitable for filamentous or colonial morphotypes that can clog flow cells and apertures. Here, we outline a new open-source package, ***cellcount***, a continuation of the report outlined in [@Pokrzywinski2019] and designed in the R computing language, to quickly and accurately count various microbial morphotypes. With this open source package, fluorescent microscope images can be analyzed to generate particle counts and quantify overall cell density from samples at a rapid speed with the accuracy of comparable conventional counting methods. Furthermore, additional tools are presented in this package using [R Shiny](https://cran.r-project.org/package=shiny) [@Chang2022] interfaces to demonstrate ease-of-use for researchers not familiar with the R environment. All materials outlined in this article, along with tutorials, can be found on NOAA's National Center for Coastal Ocean Science Harmful Algal Bloom Forecasting (NOAA NCCOS HAB-F) [GitHub.](https://github.com/orgs/HABF-SDI-NCCOS-NOAA/repositories)

@@

To get started with ***cellcount***, make a local clone of the repository using programs like [GitHub Desktop](https://desktop.github.com/). Once cloned, open the R project file and install the package locally by selecting 'build' and 'install package', or use the shortcut Crtl+Shift+B (Cmd+Shift+B for MacOS users). Once the ***cellcount*** package is installed, the following code can be used to bring ***cellcount*** into the R environment:

@@

```{r}
Cell.Count <- data.frame(Image_File_Name = character(0), Cell_Count = numeric(0))
```

@@

Once necessary adjustments have been made, the following for-loop is used to
interatively process batch images of interest. In this example, using the provided demo *Microcystis* sp.
images, the for-loop uses the *single_cell_convert* function to produce unique 
color labels on cells within an image and to produce a final image showcasing 
these labels using the *count_images* function. Finally, *count_cells* is used 
to count the number of unique color labels (i.e., cells or particles). Code has also been integrated in the for-loop
to add these count results to the 'Cell.Count' data frame made earlier, and 
to save each unique color label image to the "image_savdir" directory:

```{r}
for (z in 1:length(images)) {
  imagesConverted <- single_cell_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
  final_img <- count_images(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- count_cells(imagesConverted)
  Cell.Count[nrow(Cell.Count) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}
```

Data from the image analysis will be written to the data frame "Cell.Count", where
results can be viewed by calling the data frame:

```{r}
Cell.Count
```

@@

```{r,include=FALSE}
outlier_detect<-function(){
  mean<-median(Cell.Count$Cell_Count)
  sd<-sd(Cell.Count$Cell_Count)*2
  sd_neg<-(mean-sd)
  sd_pos<-(mean+sd)
  SD_range_detection<-ifelse(sd_pos>Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)&
    ifelse(sd_neg<Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)
  return(SD_range_detection)
}
```
```{r}
Cell.Count$Cell_Count<-as.numeric(Cell.Count$Cell_Count)
Cell.Count$SD_range<-outlier_detect()
Cell.Count
```

In order to calculate cell density of these images, the
*cell_density* function can be implemented, using equations from [@Wetzel2000]. Generating the total amount of 
cells counted from these images is needed, where it can be added to the 'Cell.Count' data frame using the following code:

```{r}
cell.total <- as.numeric(Cell.Count$Cell_Count)
cell.total <- sum(cell.total)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Total", cell.total,"null")
```

*Important note* - *the "null" data produced in the code above (and others that*
*follow) fills in the blanks from the previous data produced by the* 
***outlier_detect*** *function.*

@@

- Total counts (generated previously)
- Height of camera field of view (μm)
- Width of camera field of view (μm)
- Total number of images analyzed
- Filtration area (mm<sup>2</sup>)
- Volume filtered per image (mL)
- Total sample volume (mL)

@@

With the variables calculated, the following code uses these variables to 
calculate cell density and adds the data into the 'Cell.Count' data frame:

```{r,include=FALSE}
cell_density <- function(x, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5) {
  MCF <- filtration.area / FOV
  N <- x / images
  cyano <- (MCF * N) / volume
  cyano_total <- cyano * total.volume
  return(cyano_total)
}
```
```{r}
cell.den <- cell_density(cell.total, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Density", cell.den,"null")
```

The average number of cells per image can also be calculated, as well as cells
per mL using the following code. Note that 'cell.av' calculation must contain the
number of images the end-user has analyzed, and the 'cell.mL' must be divided
by the total volume from the original sample:

```{r}
cell.av<-(cell.total/10) #change this via total number of images
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Imaging Average",cell.av,"null")

cell.mL<-(cell.den/5) #change this via total volume
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Total Volume Cell per mL",cell.mL,"null")
```

Calling upon the 'Cell.Count' data frame can show the new generated data and to double check for any calculation errors:

```{r}
Cell.Count
```

@@

```{r}
write.csv(Cell.Count, paste0(savdir, "/F192_Test counts.csv"))
```

# Output data from test images in the ***cellcount*** package:

### Test data results for *Microcystis* strain F-192
```{r,echo=FALSE}
display(final_img)
Cell.Count
```
<br>

@@

### Test data results for *Anabena* sp.
```{r,include=FALSE}
greyscale <- function(x, contrast = 2) {
  x <- contrast * x
  x <- x[, , 1] + x[, , 2] + x[, , 3]
  x <- x / max(x)
  x <- normalize(x, inputRange = c(0.1, 0.75))
  return(x)
}
mapped <- function(x, threshold = 0.3) {
  x <- as.matrix(x)
  x[x < threshold] <- 0
  return(x)
}
border <- function(x, y) {
  c(x[1:y[1], 1], x[1:y[1], y[2]], x[1, 1:y[2]], x[y[1], 1:y[2]])
}
count_cells <- function(x) {
  countAll <- max(x)
  dims <- dim(x)
  border1 <- border(x, dims)
  ids <- unique(border1[which(border1 != 0)])
  countInner <- countAll - length(ids)
  return(countInner)
}
count_images <- function(image, normalize = TRUE, removeEdgeCells = TRUE) {
  if (removeEdgeCells) {
    dims <- dim(image)
    border1 <- border(image, dims)
    ids <- unique(border1[which(border1 != 0)])
    inner <- rmObjects(image, ids)
    EBImage::colorLabels(inner, normalize = normalize)
  } else {
    EBImage::colorLabels(image, normalize = normalize)
  }
}
single_cell_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- fillHull(image)
  image2 <- watershed(distmap(image1), tolerance = tolerance, ext = ext)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
filamentous_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- watershed(distmap(image), tolerance = tolerance, ext = ext)
  image2 <- fillHull(image1)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
cell_density <- function(x, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5) {
  MCF <- filtration.area / FOV
  N <- x / images
  cyano <- (MCF * N) / volume
  cyano_total <- cyano * total.volume
  return(cyano_total)
}
outlier_detect<-function(){
  mean<-median(Cell.Count$Cell_Count)
  sd<-sd(Cell.Count$Cell_Count)*2
  sd_neg<-(mean-sd)
  sd_pos<-(mean+sd)
  SD_range_detection<-ifelse(sd_pos>Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)&
    ifelse(sd_neg<Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)
  return(SD_range_detection)
}
```

```{r,echo=FALSE}
Cell.Count <- data.frame(Image_File_Name = character(0), Cell_Count = numeric(0))

@@

```{r,echo=FALSE}
for (z in 1:length(images)) {
  adj1<-median(img1[[z]])
  adj2<-adj1+0.2
  imagesMapped <- lapply(grey_images, mapped, threshold = adj2)
  imagesConverted <- filamentous_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
  final_img <- count_images(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- count_cells(imagesConverted)
  Cell.Count[nrow(Cell.Count) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}

Cell.Count$Cell_Count<-as.numeric(Cell.Count$Cell_Count)
Cell.Count$SD_range<-outlier_detect()

cell.total <- as.numeric(Cell.Count$Cell_Count)
cell.total <- sum(cell.total)

Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Total", cell.total,"null")

cell.den <- cell_density(cell.total, FOV = 0.027770534, images = 10, filtration.area = 213.8, volume = 0.5, total.volume =15)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Density", cell.den,"null")

cell.av<-(cell.total/10)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Imaging Average",cell.av,"null")

cell.mL<-(cell.den/15)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Total Volume Cell per mL",cell.mL,"null")

write.csv(Cell.Count, paste0(savdir, "/Anabena_test counts.csv"))

Cell.Count
```
<br>

@@

### Test data results for *Shewanella* IRI-160
```{r,include=FALSE}
greyscale <- function(x, contrast = 2) {
  x <- contrast * x
  x <- x[, , 1] + x[, , 2] + x[, , 3]
  x <- x / max(x)
  x <- normalize(x, inputRange = c(0.1, 0.75))
  return(x)
}
mapped <- function(x, threshold = 0.3) {
  x <- as.matrix(x)
  x[x < threshold] <- 0
  return(x)
}
border <- function(x, y) {
  c(x[1:y[1], 1], x[1:y[1], y[2]], x[1, 1:y[2]], x[y[1], 1:y[2]])
}
count_cells <- function(x) {
  countAll <- max(x)
  dims <- dim(x)
  border1 <- border(x, dims)
  ids <- unique(border1[which(border1 != 0)])
  countInner <- countAll - length(ids)
  return(countInner)
}
count_images <- function(image, normalize = TRUE, removeEdgeCells = TRUE) {
  if (removeEdgeCells) {
    dims <- dim(image)
    border1 <- border(image, dims)
    ids <- unique(border1[which(border1 != 0)])
    inner <- rmObjects(image, ids)
    EBImage::colorLabels(inner, normalize = normalize)
  } else {
    EBImage::colorLabels(image, normalize = normalize)
  }
}
single_cell_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- fillHull(image)
  image2 <- watershed(distmap(image1), tolerance = tolerance, ext = ext)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
filamentous_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- watershed(distmap(image), tolerance = tolerance, ext = ext)
  image2 <- fillHull(image1)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
cell_density <- function(x, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5) {
  MCF <- filtration.area / FOV
  N <- x / images
  cyano <- (MCF * N) / volume
  cyano_total <- cyano * total.volume
  return(cyano_total)
}
outlier_detect<-function(){
  mean<-median(Cell.Count$Cell_Count)
  sd<-sd(Cell.Count$Cell_Count)*2
  sd_neg<-(mean-sd)
  sd_pos<-(mean+sd)
  SD_range_detection<-ifelse(sd_pos>Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)&
    ifelse(sd_neg<Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)
  return(SD_range_detection)
}
```

```{r,echo=FALSE}
Cell.Count <- data.frame(Image_File_Name = character(0), Cell_Count = numeric(0))

@@

```{r,echo=FALSE}
for (z in 1:length(images)) {
  adj1<-median(img1[[z]])
  adj2<-adj1+0.2
  imagesMapped <- lapply(grey_images, mapped, threshold = adj2)
  imagesConverted <- single_cell_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 50, tolerance = 0.8, ext = 1)
  final_img <- count_images(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- count_cells(imagesConverted)
  Cell.Count[nrow(Cell.Count) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}

Cell.Count$Cell_Count<-as.numeric(Cell.Count$Cell_Count)
Cell.Count$SD_range<-outlier_detect()

cell.total <- as.numeric(Cell.Count$Cell_Count)
cell.total <- sum(cell.total)

Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Total", cell.total,"null")

cell.den <- cell_density(cell.total, FOV = 0.00042, images = 10, filtration.area = 213.8, volume = 0.02, total.volume =2)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Density", cell.den,"null")

cell.av<-(cell.total/10)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Imaging Average",cell.av,"null")

cell.mL<-(cell.den/2)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Total Volume Cell per mL",cell.mL,"null")

write.csv(Cell.Count, paste0(savdir, "/Shewanella_test counts.csv"))

Cell.Count
```

@@

# Package limitations:

@@

Users must be aware that this package does not contain abilities to classify distinct cell types (i.e., species). The development of this package utilized monoculture of various cyanobacteria species. Users may find difficulty in using this analysis tool if cultures contain heavy amounts of particulates, other cellular debris, and contamination of other species. It is advised to use this tool for single-species only, however future developments of this tool are planned to include identifying cell features and deep-learning mechanics that use image libraries of unique cellular species.

>>>>

The ***cellcount*** package provides a simple approach to quantifying bacteria and cyanobacteria *via* fluorescence microscopy. Microbial enumeration is an important tool for validating new methods and critical to assess potential ecological changes. However, standard manual microbial enumeration is time-consuming and notoriously difficult due to large colonies, filaments, and dense biomats. Conventional counting methods like microscopy require specialized expertise and are error prone due to variations between microscopists. Additionally, other methods of counting cells or particles (i.e. Coulter counter, flow cytometer, etc.) are not suitable for filamentous or colonial morphotypes that can clog flow cells and apertures. Here, we outline a new open-source package, ***cellcount***, a continuation of the report outlined in [@Pokrzywinski2019] and designed in the R computing language, to quickly and accurately count various microbial morphotypes. With this open source package, fluorescent microscope images can be analyzed to generate particle counts and quantify overall cell density from samples at a rapid speed with the accuracy of comparable conventional counting methods. Furthermore, additional tools are presented in this package using [R Shiny](https://cran.r-project.org/package=shiny) [@Chang2022] interfaces to demonstrate ease-of-use for researchers not familiar with the R environment. All materials outlined in this article, along with tutorials, can be found on NOAA's National Center for Coastal Ocean Science Harmful Algal Bloom Forecasting (NOAA NCCOS HAB-F) [GitHub.](https://github.com/orgs/HABF-SDI-NCCOS-NOAA/repositories)

@@

To get started with ***cellcount***, make a local clone of the repository using programs like [GitHub Desktop](https://desktop.github.com/). Once cloned, open the R project file and install the package locally by selecting 'build' and 'install package', or use the shortcut Ctrl+Shift+B (Cmd+Shift+B for MacOS users). Once the ***cellcount*** package is installed, the following code can be used to bring ***cellcount*** into the R environment:

@@

```{r}
Demo.Results <- data.frame(Image_File_Name = character(0), Cell_Count = numeric(0))
```

@@

Once necessary adjustments have been made, the following for-loop is used to
interactively process batch images of interest. In this example, using the provided demo *Microcystis* sp.
images, the for-loop uses the *single_cell_convert* function to produce unique 
color labels on cells within an image and to produce a final image showcasing 
these labels using the *count_images* function. Finally, *count_cells* is used 
to count the number of unique color labels (i.e., cells or particles). Code has also been integrated in the for-loop
to add these count results to the 'Demo.Results' data frame made earlier, and 
to save each unique color label image to the "image_savdir" directory:

```{r}
for (z in 1:length(images)) {
  imagesConverted <- single_cell_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
  final_img <- count_images(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- count_cells(imagesConverted)
  Demo.Results[nrow(Demo.Results) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}
```

Data from the image analysis will be written to the data frame 'Demo.Results, where
results can be viewed by calling the data frame:

```{r,include=FALSE}
Demo.Results
```

@@

```{r,include=FALSE}
outlier_detect<-function(){
  mean<-median(Demo.Results$Cell_Count)
  sd<-sd(Demo.Results$Cell_Count)*2
  sd_neg<-(mean-sd)
  sd_pos<-(mean+sd)
  SD_range_detection<-ifelse(sd_pos>Demo.Results$Cell_Count,Demo.Results$Cell_Count,NA)&
    ifelse(sd_neg<Demo.Results$Cell_Count,Demo.Results$Cell_Count,NA)
  return(SD_range_detection)
}
```
```{r}
Demo.Results$Cell_Count<-as.numeric(Demo.Results$Cell_Count)
Demo.Results$SD_range<-outlier_detect()
Demo.Results
```

In order to calculate cell density of these images, the
*cell_density* function can be implemented, using equations from [@Wetzel2000]. The output results need to then be converted
to a cell density variable, where it can be added to the 'Demo.Results' data frame using the following code:

```{r}
cell.total <- as.numeric(Demo.Results$Cell_Count)
cell.total <- sum(cell.total)
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Cell Total", cell.total," ")
```

@@

- Total counts (generated previously using this package)
- Height of camera field of view (μm)
- Width of camera field of view (μm)
- Total number of images analyzed
- Filtration area (mm<sup>2</sup>)
- Volume filtered per image (mL)
- Total sample volume (mL)

@@

With the variables calculated, the following code uses these variables to 
calculate cell density and adds the data into the 'Demo.Results' data frame:

```{r,include=FALSE}
cell_density <- function(x, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5) {
  MCF <- filtration.area / FOV
  N <- x / images
  cyano <- (MCF * N) / volume
  cyano_total <- cyano * total.volume
  return(cyano_total)
}
```
```{r}
cell.den <- cell_density(cell.total, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5)
cell.den <- round(cell.den,2)
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Cell Density", cell.den," ")
```

The average number of cells per image can also be calculated, as well as cells
per mL using the following code. Note that 'cell.av' calculation must contain the
number of images the end-user has analyzed (excluding poor quality images and outliers), 
and the 'cell.mL' must be divided by the total volume from the original sample where the
filtered subsample was aliquotted from:

```{r}
cell.av<-(cell.total/10) #change this via total number of images
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Imaging Average",cell.av," ")

cell.mL<-(cell.den/5) #change this via total volume
cell.mL <- round(cell.mL,2)
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Total Volume Cell per mL",cell.mL," ")
```

Calling upon the 'Demo.Results' data frame can show the new generated data and to double check for any calculation errors:

```{r,include=FALSE}
Demo.Results
```

@@

```{r}
write.csv(Demo.Results, paste0(savdir, "/F192_Test counts.csv"))
```

# Output data from test images in 
#the ***cellcount*** package:

### Test data results for *Microcystis* strain F-192
```{r,echo=FALSE}
display(final_img)
Demo.Results
```
<br>

@@

### Test data results for *Anabaena* sp.
```{r,include=FALSE}
greyscale <- function(x, contrast = 2) {
  x <- contrast * x
  x <- x[, , 1] + x[, , 2] + x[, , 3]
  x <- x / max(x)
  x <- normalize(x, inputRange = c(0.1, 0.75))
  return(x)
}
mapped <- function(x, threshold = 0.3) {
  x <- as.matrix(x)
  x[x < threshold] <- 0
  return(x)
}
border <- function(x, y) {
  c(x[1:y[1], 1], x[1:y[1], y[2]], x[1, 1:y[2]], x[y[1], 1:y[2]])
}
count_cells <- function(x) {
  countAll <- max(x)
  dims <- dim(x)
  border1 <- border(x, dims)
  ids <- unique(border1[which(border1 != 0)])
  countInner <- countAll - length(ids)
  return(countInner)
}
count_images <- function(image, normalize = TRUE, removeEdgeCells = TRUE) {
  if (removeEdgeCells) {
    dims <- dim(image)
    border1 <- border(image, dims)
    ids <- unique(border1[which(border1 != 0)])
    inner <- rmObjects(image, ids)
    EBImage::colorLabels(inner, normalize = normalize)
  } else {
    EBImage::colorLabels(image, normalize = normalize)
  }
}
single_cell_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- fillHull(image)
  image2 <- watershed(distmap(image1), tolerance = tolerance, ext = ext)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
filamentous_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- watershed(distmap(image), tolerance = tolerance, ext = ext)
  image2 <- fillHull(image1)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
cell_density <- function(x, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5) {
  MCF <- filtration.area / FOV
  N <- x / images
  cyano <- (MCF * N) / volume
  cyano_total <- cyano * total.volume
  return(cyano_total)
}
outlier_detect<-function(){
  mean<-median(Demo.Results$Cell_Count)
  sd<-sd(Demo.Results$Cell_Count)*2
  sd_neg<-(mean-sd)
  sd_pos<-(mean+sd)
  SD_range_detection<-ifelse(sd_pos>Demo.Results$Cell_Count,Demo.Results$Cell_Count,NA)&
    ifelse(sd_neg<Demo.Results$Cell_Count,Demo.Results$Cell_Count,NA)
  return(SD_range_detection)
}
```

```{r,echo=FALSE}
Demo.Results <- data.frame(Image_File_Name = character(0), Cell_Count = numeric(0))

@@

```{r,echo=FALSE}
for (z in 1:length(images)) {
  adj1<-median(img1[[z]])
  adj2<-adj1+0.2
  imagesMapped <- lapply(grey_images, mapped, threshold = adj2)
  imagesConverted <- filamentous_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
  final_img <- count_images(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- count_cells(imagesConverted)
  Demo.Results[nrow(Demo.Results) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}

Demo.Results$Cell_Count<-as.numeric(Demo.Results$Cell_Count)
Demo.Results$SD_range<-outlier_detect()

cell.total <- as.numeric(Demo.Results$Cell_Count)
cell.total <- sum(cell.total)

Demo.Results[nrow(Demo.Results) + 1, ] <- c("Cell Total", cell.total," ")

cell.den <- cell_density(cell.total, FOV = 0.027770534, images = 10, filtration.area = 213.8, volume = 0.5, total.volume =15)
cell.den <- round(cell.den,2)
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Cell Density", cell.den," ")

cell.av<-(cell.total/10)
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Imaging Average",cell.av," ")

cell.mL<-(cell.den/15)
cell.mL <- round(cell.mL,2)
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Total Volume Cell per mL",cell.mL," ")

write.csv(Demo.Results, paste0(savdir, "/Anabena_test counts.csv"))

Demo.Results
```
<br>

@@

### Test data results for *Shewanella* IRI-160
```{r,include=FALSE}
greyscale <- function(x, contrast = 2) {
  x <- contrast * x
  x <- x[, , 1] + x[, , 2] + x[, , 3]
  x <- x / max(x)
  x <- normalize(x, inputRange = c(0.1, 0.75))
  return(x)
}
mapped <- function(x, threshold = 0.3) {
  x <- as.matrix(x)
  x[x < threshold] <- 0
  return(x)
}
border <- function(x, y) {
  c(x[1:y[1], 1], x[1:y[1], y[2]], x[1, 1:y[2]], x[y[1], 1:y[2]])
}
count_cells <- function(x) {
  countAll <- max(x)
  dims <- dim(x)
  border1 <- border(x, dims)
  ids <- unique(border1[which(border1 != 0)])
  countInner <- countAll - length(ids)
  return(countInner)
}
count_images <- function(image, normalize = TRUE, removeEdgeCells = TRUE) {
  if (removeEdgeCells) {
    dims <- dim(image)
    border1 <- border(image, dims)
    ids <- unique(border1[which(border1 != 0)])
    inner <- rmObjects(image, ids)
    EBImage::colorLabels(inner, normalize = normalize)
  } else {
    EBImage::colorLabels(image, normalize = normalize)
  }
}
single_cell_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- fillHull(image)
  image2 <- watershed(distmap(image1), tolerance = tolerance, ext = ext)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
filamentous_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- watershed(distmap(image), tolerance = tolerance, ext = ext)
  image2 <- fillHull(image1)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
cell_density <- function(x, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5) {
  MCF <- filtration.area / FOV
  N <- x / images
  cyano <- (MCF * N) / volume
  cyano_total <- cyano * total.volume
  return(cyano_total)
}
outlier_detect<-function(){
  mean<-median(Demo.Results$Cell_Count)
  sd<-sd(Demo.Results$Cell_Count)*2
  sd_neg<-(mean-sd)
  sd_pos<-(mean+sd)
  SD_range_detection<-ifelse(sd_pos>Demo.Results$Cell_Count,Demo.Results$Cell_Count,NA)&
    ifelse(sd_neg<Demo.Results$Cell_Count,Demo.Results$Cell_Count,NA)
  return(SD_range_detection)
}
```

```{r,echo=FALSE}
Demo.Results <- data.frame(Image_File_Name = character(0), Cell_Count = numeric(0))

@@

```{r,echo=FALSE}
for (z in 1:length(images)) {
  adj1<-median(img1[[z]])
  adj2<-adj1+0.2
  imagesMapped <- lapply(grey_images, mapped, threshold = adj2)
  imagesConverted <- single_cell_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 50, tolerance = 0.8, ext = 1)
  final_img <- count_images(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- count_cells(imagesConverted)
  Demo.Results[nrow(Demo.Results) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}

Demo.Results$Cell_Count<-as.numeric(Demo.Results$Cell_Count)
Demo.Results$SD_range<-outlier_detect()

cell.total <- as.numeric(Demo.Results$Cell_Count)
cell.total <- sum(cell.total)

Demo.Results[nrow(Demo.Results) + 1, ] <- c("Cell Total", cell.total," ")

cell.den <- cell_density(cell.total, FOV = 0.00042, images = 10, filtration.area = 213.8, volume = 0.02, total.volume =2)
cell.den <- round(cell.den,2)
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Cell Density", cell.den," ")

cell.av<-(cell.total/10)
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Imaging Average",cell.av," ")

cell.mL<-(cell.den/2)
cell.mL <- round(cell.mL,2)
Demo.Results[nrow(Demo.Results) + 1, ] <- c("Total Volume Cell per mL",cell.mL," ")

write.csv(Demo.Results, paste0(savdir, "/Shewanella_test counts.csv"))

Demo.Results
```

@@

# Package limitations/constraints:

@@

Users must be aware that this package does not contain abilities to classify distinct cell types (i.e., species). The development of this package utilized mono-culture of various cyanobacteria species. Users may find difficulty in using this analysis tool if cultures contain heavy amounts of particulates, other cellular debris, and contamination of other species. It is advised to use this tool for single-species only, however future developments of this tool are planned to include identifying cell features and deep-learning mechanics that use image libraries of unique cellular species.
